---
title: "ProVoC"
format: gfm
execute:
  cache: true
---

[![Lifecycle:
development](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental-1)
[![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/licenses/MIT)

PROportions of Variants of Concern using counts, coverage, and a variant matrix.

Builds and diagnoses a model based on:

- Counts: The number of times a given mutation was observed.
- Coverage: The number of times the position of a given variant was read.
- Mutation names: whatever format you want, as long as they match the names in the Variant Matrix.
- Variant Matrix: A matrix where the row names are variants, the column names are mutations, and the entries are 1 if the row variant has the column mutation and 0 otherwise. 
    - For example: Usher Barcodes
    - All current methods accept fractional entries.

# Usage

There are two steps to using this software: create the mutation definitions and run the model(s).

If not specified, the mutation definitions uses hardcoded definitions from the [cov-lineages/constellations](https://github.com/cov-lineages/constellations) repo, which contains the representative mutations that were identified by the PANGO team. The mutation definitions must have names that match what exists in the data.

```{r}
#| label: load_Baaijens
#| echo: true
#| eval: true
library(provoc)
data(Baaijens)
b1 <- Baaijens [Baaijens$sra == Baaijens$sra[1], ]
b1$mutation <- parse_mutations(b1$label)

head(b1[, c("count", "coverage", "mutation", "label")])
```

The main fitting function is designed to mimic `glm()`, with formula notation that emphasizes the connection to binomial GLM models. The `bootstrap_samples` argument is optional, and is used to calculate confidence intervals and correlation of the parameters.

```{r}
#| label: fit one sample
res <- provoc(
    formula = count / coverage ~ B.1.1.7 + B.1.429 + B.1.617.2 + 
        B.1.427 + AY.4 + P.1,
    data = b1,
    bootstrap_samples = 100)
res
```

We have created a class for `provoc` objects with convenient methods. For example, plotting the results is achieved as follows:

```{r}
#| label: one-sample-res-base
#| fig-height: 4
plot(res)
```

We use the convention of [`autoplot()`](https://ggplot2.tidyverse.org/reference/autoplot.html) as a function that creates a `ggplot2` plot based on a particular type of data. This allows for all of `ggplot2`'s fanciness on top of a pre-made plot.

```{r}
#| label: one-sample-res-ggplot
#| fig-height: 4
library(ggplot2)
autoplot(res) +
    theme_bw() +
    labs(title = "Results for one sample")
```

```{r}
#| label: one-sample-resid-plot
#| fig-height: 4
plot_resids(res)
summary(res)
```

```{r}
#| label: one-sample-variant-plot
plot_variants(res)
```

# Multiple Samples

```{r}
#| label: multi_sample
# First two samples from Baaijens
b2 <- Baaijens [Baaijens$sra %in% unique(Baaijens$sra)[1:30], ]
b2$mutations <- parse_mutations(b2$label)
head(b1[, c("count", "coverage", "mutation", "label", "sra")])
```

Note the "`by`" argument below. This tells `provoc()` to fit the model separately to each sample defined by the `by` column. Since there are 30 samples, the model takes a while to run, hence why there are only 50 bootstrap samples here.

```{r}
#| label: fit-multi-sample
#| warning: false
#| message: false
system.time(
    res <- provoc(
        formula = count / coverage ~ B.1.1.7 + B.1.429 + B.1.617.2 +
            B.1.427 + AY.4 + P.1,
        data = b2, 
        by = "sra",
        bootstrap_samples = 0)
)
res
```


```{r}
#| label: multi-sample-variant-plot
plot_variants(res)
```


The plotting functions above work as expected.

```{r}
#| fig-height: 8
#| label: multi-sample-res-base
plot(res)
```


```{r}
#| fig-height: 8
#| label: multi-sample-resid-base
plot_resids(res)
```



```{r}
#| fig-height: 4
#| label: multi-sample-res-ggplot
autoplot(res)
```

In addition, we've added special functionality for data with a "date" column:

```{r}
#| fig-height: 4
#| label: multi-date-res-ggplot
theme_set(theme_bw())
res$date <- lubridate::ymd(res$date)
autoplot(res, date_col = "date")
```


```{r}
#| label: multi-sample-resid-plot
#| fig-height: 4
plot_resids(res)
```

## Searching for Different Variants

```{r}
#| label: with-without
library(patchwork)
res_with <- provoc(cbind(count, coverage) ~ B.1.1.7 + B.1.429 + B.1.617.2 + B.1.427,
    data = b2, by = "sra",
    verbose = FALSE, bootstrap_samples = 0)
res_with$date <- lubridate::ymd(res_with$date)

res_without <- provoc(cbind(count, coverage) ~ B.1.429 + B.1.617.2 + B.1.427,
    data = b2, by = "sra",
    verbose = FALSE, bootstrap_samples = 0)
res_without$date <- lubridate::ymd(res_without$date)

g_with <- autoplot(res_with, date_col = "date") +
    scale_fill_manual(values = 2:5)
g_without <- autoplot(res_without, date_col = "date") +
    scale_fill_manual(values = 3:5)

g_with / g_without
```

# Roadmap

- Soon
    - [ ] Finish simplified versions of Freyja and Alcov models.
    - [ ] Full suite of model diagnostic visualizations.
    - [ ] Process data from iVar for use in models.
        - [ ] Process a list or directory of iVar data into an appropriate object.
    - [ ] Tools to investigate the effect of different lineage definitions.
- Medium future
    - [ ] Convert Freyja and Alcov outputs to R objects that can take advantage of the diagnostics.
    - [ ] Allow lists of formulae, lists of mutation_defs, and lists of data.
- Far future
    - [ ] Make available on CRAN
