---
title: "ProVoC"
format: gfm
execute:
  cache: true
---

[![Lifecycle: development](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental-1) [![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/licenses/MIT)

PROportions of Variants of Concern using counts, coverage, and a lineage definition matrix.

Builds and diagnoses a model based on:

- Counts: The number of times a given mutation was observed.
- Coverage: The number of times the position of a given mutation was read.
- Mutation names: whatever format you want, as long as they match the names in the lineage definition matrix.
- Lineage definition matrix: A matrix where the row names are lineage, the column names are mutations, and the entries are 1 if the row lineage has the column mutation and 0 otherwise.
    - For example: Usher Barcodes
    - All current methods accept fractional entries.

# Usage

There are two steps to using this software: create the lineage definitions and run the model(s).

If not specified, the lineage definitions uses hardcoded definitions from the [cov-lineages/constellations](https://github.com/cov-lineages/constellations) repo, which contains the representative mutations that were identified by the PANGO team. The lineage definitions must have names that match what exists in the data.

```{r}
#| label: load_Baaijens
#| echo: true
#| eval: true
#library(provoc)
devtools::load_all()
data(Baaijens)
b1 <- Baaijens [Baaijens$sra == Baaijens$sra[1], ]
b1$mutation <- parse_mutations(b1$label)

head(b1[, c("count", "coverage", "mutation", "label")])
```

The main fitting function is designed to mimic `glm()`, with formula notation that emphasizes the connection to binomial GLM models. The `bootstrap_samples` argument is optional, and is used to calculate confidence intervals and correlation of the parameters.

```{r}
#| label: fit one sample
res <- provoc(
    formula = count / coverage ~ B.1.1.7 + B.1.429 + B.1.617.2 + 
        B.1.427 + AY.4 + P.1,
    data = b1,
    bootstrap_samples = 100)
res
```

We have created a class for `provoc` objects with convenient methods. For example, plotting the results is achieved as follows:

```{r}
#| label: one-sample-res-base
#| fig-height: 4
plot(res)
```

We use the convention of [`autoplot()`](https://ggplot2.tidyverse.org/reference/autoplot.html) as a function that creates a `ggplot2` plot based on a particular type of data. This allows for all of `ggplot2`'s fanciness on top of a pre-made plot.

```{r}
#| label: one-sample-res-ggplot
#| fig-height: 4
library(ggplot2)
autoplot(res) +
    theme_bw() +
    labs(title = "Results for one sample")
```

```{r}
#| label: one-sample-resid-plot
#| fig-height: 4
plot_resids(res)
summary(res)
```

```{r}
#| label: one-sample-lineage-plot
plot_lineages(res)
```

# Multiple Samples

```{r}
#| label: multi_sample
# First thirty samples from Baaijens
b2 <- Baaijens [Baaijens$sra %in% unique(Baaijens$sra)[1:30], ]
b2$mutations <- parse_mutations(b2$label)
head(b1[, c("count", "coverage", "mutation", "label", "sra")])
```

Note the "`by`" argument below. This tells `provoc()` to fit the model separately to each sample defined by the `by` column. Since there are 30 samples, the model takes a while to run, hence why there are no bootstrap samples here.

```{r}
#| label: fit-multi-sample
#| warning: false
#| message: false
system.time(
    res <- provoc(
        formula = count / coverage ~ B.1.1.7 + B.1.429 + B.1.617.2 +
            B.1.427 + AY.4 + P.1,
        data = b2, 
        by = "sra",
        bootstrap_samples = 0)
)
res
```

```{r}
#| label: multi-sample-lineage-plot
plot_lineages(res)
```

The plotting functions above work as expected.

```{r}
#| fig-height: 8
#| label: multi-sample-res-base
plot(res)
```

```{r}
#| fig-height: 8
#| label: multi-sample-resid-base
plot_resids(res)
```

```{r}
#| fig-height: 4
#| label: multi-sample-res-ggplot
autoplot(res)
```

In addition, we've added special functionality for data with a "date" column:

```{r}
#| fig-height: 4
#| label: multi-date-res-ggplot
theme_set(theme_bw())
res$date <- lubridate::ymd(res$date)
autoplot(res, date_col = "date")
```

Since `autoplot()` returns a `ggplot2` object, you can use all your knowledge of ggplot2:

```{r}
#| label: faceting
autoplot(res, date_col = "date") +
    facet_wrap(~ lineage, nrow = 1) +
    labs(title = "Proportions of Lineages over Time") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Finally, the residual plot still works (although it's a bunch of samples stuffed into one plot, so its use is limited.)

```{r}
#| label: multi-sample-resid-plot
#| fig-height: 4
plot_resids(res)
```

## Searching for Different Lineages

```{r}
#| label: with-without
library(patchwork)
res_with <- provoc(cbind(count, coverage) ~ B.1.1.7 + B.1.429 + B.1.617.2 + B.1.427,
    data = b2, by = "sra",
    verbose = FALSE, bootstrap_samples = 0)
res_with$date <- lubridate::ymd(res_with$date)

res_without <- provoc(cbind(count, coverage) ~ B.1.429 + B.1.617.2 + B.1.427,
    data = b2, by = "sra",
    verbose = FALSE, bootstrap_samples = 0)
res_without$date <- lubridate::ymd(res_without$date)

g_with <- autoplot(res_with, date_col = "date") +
    scale_fill_manual(values = 2:5)
g_without <- autoplot(res_without, date_col = "date") +
    scale_fill_manual(values = 3:5)

g_with / g_without
```

Notice two important points:

1. Since proportions can't sum to greater than one, including B.1.1.7 when it should be excluded means it will take away from the other lineages!
    - Erroneously **increasing** one means **decreasing** true lineages.
2. Due to shared mutations, excluding B.1.1.7 when it should be included means other lineages increase to make up for it.
    - Erroneously **decreasing** one means **increasing** other lineages.

`provoc` also accepts the `~ .` formula notation. This is useful if you want to manipulate the lineages in the lineage definitions matrix, which is easier programmatically than dealing with formulae in R. The `filter_lineages()` function will remove mutations that are not present in any of the remaining lineages.

```{r}
#| label: load-usher
lineage_defs <- provoc::usher_barcodes(path = "working") |>
    filter_lineages(c("B.1.1.7", "B.1.617.2", "B.1.427",
        "B.1.429", "AY.4", "BA.4", "BA.5"))
```

With these definitions, we can just use all of them:

```{r}
#| label: run-barcodes
res_all <- provoc(count / coverage ~ ., data = b1, 
    bootstrap_samples = 0,
    lineage_defs = lineage_defs)

plot_lineages(res_all)

```

# Version Roadmap

- [x] 0.5: Current version
- [ ] 0.6: Refactor and add methods
    - [ ] Standardize to use "lineage" everywhere
    - [ ] General ease of use for fitting multiple models, checking lineage definitions, and checking diagnostics.
    - [ ] Correct documentation errors and vestiges
- [ ] 0.7: Better handling of multiple samples
    - [ ] Proper diagnostics, easily applied to all samples
    - [ ] `multiprovoc` class so functions know how to handle things?
- [ ] 0.8: Multiple methods
    - [ ] Freyja and AlCoV (possibly others) as first-class objects, with access to methods for `provoc` objects.
        - Potentially loading in Freyja output files as `provoc` objects
    - [ ] Stacking (`cbind()`) models for ease of comparison
    - [ ] Import and process iVar data.
    - [ ] Main function works for multiple methods and data sources.
    - [ ] Unified `parse_mutations()` function with minimal user input.
- [ ] 0.9: The shiny minor version
    - [ ] Shiny app to interact with lineage defitions
    - [ ] Shiny app to interact with results
- [ ] 1.0: First major release
    - [ ] All functions unit-tested and debugged
    - [ ] Vignettes fully built
    - [ ] All documentation up-to-date and built via `pkgdown`
    - [ ] 
