---
title: "building-mutation-lists"
output: rmarkdown::html_vignette
vignette: >
    %\VignetteIndexEntry{building-mutation-lists}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

```{r setup}
library(provoc)
```

# Overview

There are three ways to build mutation lists provided by ProVoC:

1. Using a mutation list curated by the PANGO team ("Constellations")
    - These are (mostly) representative mutations - there is very little overlap between the mutation lists for different lineages.
    - The team takes great pains to keep them updated and useful.
    - This list only a specific subset of lineages of concern with a small subset of their mutations and may not be useful for all purposes.
2. By specifying the lineages that are of interest to you and then pruning down the NextStrain GenBank data.
3. Using lineages that contain the same mutations as the ones in your data.

# Constellations

The PANGO team maintains a list of (mostly) representative mutations for Variants of Concern in the github repo [cov-lineages/constellations](https://github.com/cov-lineages/constellations).
ProVoC has a built-in function to pull from these lists, but relies on the user to keep the list up to date using git commands.
ProVoC expects to be in the base folder of your project directory and the constellations repo to be in the same parent directory, i.e.

```
git-projects/
├── wastewater-voc/
│   ├── folders
│   └── stuff
└── constellations/
    └── rest-of-constellations-rep
```

```sh
ll # check that your project directory is in the current directory
git clone https://github.com/cov-lineages/constellations
```

From within R, you can get the lineage definition matrix using the following code:

```{r}
lineage_defs <- astronomize()
```

Note that this structure isn't required.
The `astronomize()` function takes a `path` argument.
For example, the `here()` package always finds your root directory, so if your constellations repo is in the `~/other/constellations` folder and your project directory is in `~/research/wastewater`, then the R code `astronomize(here("../../other/constellations"))` will work.
This is a lot of work to say that you should have a `git` folder with all your projects in one place, including the constellations repo.

Thanks to the magic of version control, you can go back in time to previous versions of the constellations repo.
This allows you to build mutation lists based on what was known at the time.
This is useful to ensure that you're not looking for BA.4 in samples from early 2021, for instance.
Running the following git command in a shell (e.g. Konsole, Terminal, or Git Bash) in the constellations repo will revert it to the most recent commit as of 2021-12-01.

```sh
git checkout `git rev-list -n 1 --first-parent --before="2021-12-01 00:01" master`
```

The repo can be brought back up to date with a simple `git checkout`.



# Specifying Lineages

ProVoC also contains a function to build mutation lists based on a list of lineage names.
It does so using the NextStrain GenBank metadata, which contains a list of mutations relative to the Wuhan-1 genome sequence.

The function `lineage_defs_by_lineages()` allows for a specification of the sample date.
The mutations in the `mutations_by_lineage` data frame include information about the dates that the mutation was observed.
For instance, a single sequence assigned to the lineage B.1.1.7 was observed in the summer of 2020, even though this lineage only became a variant of concern in the spring of 2021.
The lineage had acquired many new mutations by the time it became a variant of concern, and it would thus be inappropriate to assume that wastewater sampled in summer of 2020 would be expected to contain the new mutations.

The data set is separated by lineage, which each lineage having one row dedicated to each mutation observed in that lineage.
There is a column representing the first time a given mutation was observed in that lineage, and a column representing the last time that mutation was observed in that lineage.

The function also incorporates the date that the database  to ensure that currently circulating lineages are not assumed to be missing simply because they haven't been sequenced recently.
If the last observed date is within the last two months since the GenBank data were accessed, it is assumed that that mutation is still present in that lineage.



